---
title: "lab5"
author: "Michelle H"
format:
  html:
    embed-resources: true
editor: visual
---

```{r}
library(tidyverse)
library(data.table)
library(dtplyr)
library(dplyr)
```

## load the data

```{r}
if (!file.exists("met_all.gz"))
  download.file(
    url = "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz",
    destfile = "met_all.gz",
    method   = "libcurl",
    timeout  = 60
  )
met <- fread("met_all.gz")
```

```{r}
# Download the data
stations <- fread("ftp://ftp.ncdc.noaa.gov/pub/data/noaa/isd-history.csv")
stations[, USAF := as.integer(USAF)]


stations[, USAF   := fifelse(USAF == 999999, NA_integer_, USAF)]
stations[, CTRY   := fifelse(CTRY == "", NA_character_, CTRY)]
stations[, STATE  := fifelse(STATE == "", NA_character_, STATE)]


stations <- unique(stations[, list(USAF, CTRY, STATE)])


stations <- stations[!is.na(USAF)]


stations[, n := 1:.N, by = .(USAF)]
stations <- stations[n == 1,][, n := NULL]
```

### Merge the data

```{r}

dat <- merge(
  # Data
  x     = met,      
  y     = stations, 
  by.x  = "USAFID",
  by.y  = "USAF", 
  all.x = TRUE,      
  all.y = FALSE
  )
head(dat[, list(USAFID, WBAN, STATE)], n = 4)
```

## Q.1 **Representative station for the US**

```{r}
dat[, (median(temp, na.rm = TRUE)), by= .(USAFID)]
dat[median(temp, na.rm = TRUE)]
median_wind <- dat[median(wind.sp, na.rm = TRUE)]
dat[median(atm.press, na.rm = TRUE)]
```

```{r}
#medians

#wind
median_wind_site <- dat[median(wind.sp, na.rm = TRUE)]

wind_quantiles <- quantile(dat$wind.sp, na.rm = TRUE)
median_wind <- median(dat$wind.sp, na.rm = TRUE)
median_station_wind <- dat$USAFID[which.min(abs(dat$wind.sp - median_wind))]

median_wind_site # 690150 
wind_quantiles #  2.1
median_wind #2.1
median_station_wind #720113

#temp
median_temp_site <- dat[median(temp, na.rm = TRUE)]

temp_quantiles <- quantile(dat$temp, na.rm = TRUE)

median_temp <- median(dat$temp, na.rm = TRUE)
median_station_temp <- dat$USAFID[which.min(abs(dat$temp - median_temp))]

median_temp_site # 690150
temp_quantiles # 23.5
median_temp # 23.5
median_station_temp # 720113

#atm pressure
median_atm_site <- dat[median(atm.press, na.rm = TRUE)]

atm_quantiles <- quantile(dat$atm.press, na.rm = TRUE)
median_atm <- median(dat$atm.press, na.rm = TRUE)
median_station_atm <- dat$USAFID[which.min(abs(dat$atm.press - median_atm))]

median_atm_site # 720110
atm_quantiles # 1014.1
median_atm # 1014.1
median_station_atm #690150                      

```

The median station does not coincide with the quantile method.

## Q.2 **Representative station per state**

Just like the previous question, you are asked to identify what is the most representative, the median, station per state. This time, instead of looking at one variable at a time, look at the euclidean distance. If multiple stations show in the median, select the one located at the lowest latitude.

Knit the doc and save it on GitHub.

```{r}

#  ## WORKING
library(dplyr)

med.temp <- median(dat$temp, na.rm = TRUE)
med.wind.sp <- median(dat$wind.sp, na.rm = TRUE)
med.atm.press <- median(dat$atm.press, na.rm = TRUE)

station_distances <- dat %>%
  mutate(
    temp_distance = ifelse(is.na(temp), NA, abs(temp - med.temp)),
    wind_distance = ifelse(is.na(wind.sp), NA, abs(wind.sp - med.wind.sp)),
    atm_distance = ifelse(is.na(atm.press), NA, abs(atm.press - med.atm.press))
  ) %>%
  group_by(STATE) %>%
  summarise(
    USAFID = USAFID[which.min(temp_distance + wind_distance + atm_distance)],
    temp = temp[which.min(temp_distance + wind_distance + atm_distance)],
    wind.sp = wind.sp[which.min(temp_distance + wind_distance + atm_distance)],
    atm.press = atm.press[which.min(temp_distance + wind_distance + atm_distance)]
  ) %>%
  ungroup()

station_distances

```

## **Question 3: In the middle?**

For each state, identify what is the station that is closest to the mid-point of the state. Combining these with the stations you identified in the previous question, use `leaflet()` to visualize all \~100 points in the same figure, applying different colors for those identified in this question.

Knit the doc and save it on GitHub.

```{r}
med.lon <- sapply(split(dat$lon, dat$STATE), median, na.rm = TRUE)
med.lat <- sapply(split(dat$lat, dat$STATE), median, na.rm = TRUE)

median_pt <- data.table(
  state = names(med.lon),
  longitude = med.lon,
  latitude = med.lat
)

station_center <- dat %>%
  mutate(
    lon_distance = ifelse(is.na(lon), NA, abs(lon - med.lon)),
    lat_distance = ifelse(is.na(lat), NA, abs(lat - med.lat))) %>%
  group_by(STATE) %>%
  summarise(
    USAFID = USAFID[which.min(lon_distance + lat_distance)],
    lon = lon[which.min(lon_distance + lat_distance)],
    lat = lat[which.min(lon_distance + lat_distance)],
  ) %>%
  ungroup()

station_center
```

```{r}
library(leaflet)
library(dplyr)

euc_map <- leaflet() %>%
  addProviderTiles('CartoDB.Positron') 

euc_map <- euc_map %>%
  addCircleMarkers(
    data = median_pt,
    lng = ~longitude,
    lat = ~latitude,
    color = "blue",
    radius = 6,
    popup = ~state
  )

euc_map <- euc_map %>%
  addCircleMarkers(
    data = station_center,
    lng = ~lon,
    lat = ~lat,
    color = "red",
    radius = 6,
    popup = ~STATE
  )


euc_map

```

## Q.4 **Means of means**

Using the `quantile()` function, generate a summary table that shows the number of states included, average temperature, wind-speed, and atmospheric pressure by the variable "average temperature level," which you'll need to create.

Start by computing the states' average temperature. Use that measurement to classify them according to the following criteria:

-   low: temp \< 20

-   Mid: temp \>= 20 and temp \< 25

-   High: temp \>= 25

Once you are done with that, you can compute the following:

-   Number of entries (records),

-   Number of NA entries,

-   Number of stations,

-   Number of states included, and

-   Mean temperature, wind-speed, and atmospheric pressure.

```{r}
test_temp <- sapply(split(dat$temp, dat$STATE), quantile, na.rm = TRUE)
quantiles_50 <- test_temp["50%",]
quantiles_50
```

```{r}
dat2 <- dat %>%
  group_by(STATE) %>%
  mutate(Median_Temperature = median(temp, na.rm = TRUE)) %>%
  ungroup()
dat2


dat3 <- dat2 %>%
  group_by(STATE) %>%
  mutate(Median_Temperature = median(temp, na.rm = TRUE),
         temp_cat = case_when(
           Median_Temperature < 20 ~ 'Low temp',
           Median_Temperature >= 20 & Median_Temperature < 25 ~ 'Mid temp',
           Median_Temperature >= 25 ~ 'High temp',
           TRUE ~ 'NA'
         )) %>%
  ungroup()
dat3

sum(dat3$temp_cat=='Low temp')
sum(dat3$temp_cat=='Mid temp')
sum(dat3$temp_cat=='High temp')
sum(430794,1135423,811126 )
```

### Low Temp

```{r}
#Number of entries (records),
sum(dat3$temp_cat=='Low temp')

#Number of NA entries,
sum(dat3$temp_cat=='NA')
length(is.na(dat3$temp_cat=='Low temp'))

#Number of stations,
unique(dat3$USAFID[dat3$temp_cat == 'Low temp'])

sum(unique(dat3$USAFID[dat3$temp_cat == 'Low temp']))


#Number of states included, and
unique(dat3$STATE[dat3$temp_cat == 'Low temp'])
length(unique(dat3$STATE[dat3$temp_cat == 'Low temp']))
#11

#Mean temperature, wind-speed, and atmospheric pressure.
mean(dat3$temp[dat3$temp_cat == 'Low temp'], na.rm = TRUE)
#18.96446

```

### Mid Temp

```{r}
#Number of entries (records),
sum(dat3$temp_cat=='Mid temp')

#Number of NA entries,
sum(dat3$temp_cat=='NA')

#Number of stations,
unique(dat3$USAFID[dat3$temp_cat == 'Mid temp'])

sum(unique(dat3$USAFID[dat3$temp_cat == 'Mid temp']))
#565356157

#Number of states included, and
unique(dat3$STATE[dat3$temp_cat == 'Mid temp'])
length(unique(dat3$STATE[dat3$temp_cat == 'Mid temp']))
#25

#Mean temperature, wind-speed, and atmospheric pressure.
mean(dat3$temp[dat3$temp_cat == 'Mid temp'], na.rm = TRUE)
#22.39909

```

### High Temp

```{r}
#Number of entries (records),
sum(dat3$temp_cat=='High temp')

#Number of NA entries,
sum(dat3$temp_cat=='NA')

#Number of stations,
unique(dat3$USAFID[dat3$temp_cat == 'High temp'])

sum(unique(dat3$USAFID[dat3$temp_cat == 'High temp']))
#400804953

#Number of states included, and
unique(dat3$STATE[dat3$temp_cat == 'High temp'])
length(unique(dat3$STATE[dat3$temp_cat == 'High temp']))
#12

#Mean temperature, wind-speed, and atmospheric pressure.
mean(dat3$temp[dat3$temp_cat == 'High temp'], na.rm = TRUE)
#27.75066
```
